---
title: "Introduction: Adjusted Predictions and Marginal Effects for Random Effects Models"
author: "Daniel Lüdecke"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction: Adjusted Predictions and Marginal Effects for Random Effects Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r set-options, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", dev = "png", fig.width = 7, fig.height = 4, message = FALSE, warning = FALSE)
options(width = 800)
dev_version <- length(strsplit(packageDescription("ggeffects")$Version, ".", fixed = TRUE)[[1]]) > 3
if (!requireNamespace("ggplot2", quietly = TRUE) ||
    !requireNamespace("lme4", quietly = TRUE) ||
    !requireNamespace("glmmTMB", quietly = TRUE) ||
    !requireNamespace("sjlabelled", quietly = TRUE)) {
  knitr::opts_chunk$set(eval = FALSE)
}
```

This vignette shows how to calculate marginal effects that take the random-effect variances for mixed models into account. 

## Marginal effects for mixed effects models

Basically, the type of predictions, i.e. whether to account for the uncertainty of random effects or not, can be set with the `type`-argument. The default, `type = "fixed"`, means that predictions are on the population-level and do not account for the random effect variances. Intervals are _confidence intervals_ for the predicted values.

```{r}
library(ggeffects)
library(lme4)
data(sleepstudy)
m <- lmer(Reaction ~ Days + (1 + Days | Subject), data = sleepstudy)

pr <- ggpredict(m, "Days")
pr
plot(pr)
```

### Prediction intervals

When `type = "random"`, the predicted _values_ are still on the _population-level_, but conditioned on the random-effect variances (individual level predictions), which means that `re.form = NULL` when calling `predict()`. However, this _may_ affect the returned predicted values, depending on whether `REML = TRUE` or `REML = FALSE` was used for model fitting. Furthermore, when `type = "random"`, the random-effect _variances_ are taken into account, meaning that the intervals are actually _prediction intervals_ and become larger. More technically speaking, `type = "random"` accounts for the uncertainty of the fixed effects _conditional on the estimates_ of the random-effect variances and conditional modes (BLUPs).

Predicted values are _not_ on the population-level when a particular group (from the random effects) is specified in the `condition` or `terms` arguments. See examples below (section _Marginal effects for each level of random effects_) for details.

The random-effect variance is the _mean_ random-effect variance. Calculation is based on the proposal from _Johnson et al. 2014_, which is also implemented in functions like [`performance::r2()`](https://easystats.github.io/performance/reference/r2_nakagawa.html) or [`insight::get_variance()`](https://easystats.github.io/insight/reference/get_variance.html) to get r-squared values or random-effect variances for mixed models with more complex random effects structures.

As can be seen, compared to the previous example with `type = "fixed"`, predicted values are identical (both on the population-level). However, standard errors, and thus the resulting confidence (or prediction) intervals are much larger .

```{r}
pr <- ggpredict(m, "Days", type = "random")
pr
plot(pr)
```

### Conditional predictions at the population-level

The reason why both `type = "fixed"` and `type = "random"` return predictions at population-level is because `ggpredict()` returns predicted values of the response _at specific levels_ of given model predictors, which are defined in the data frame that is passed to the `newdata`-argument (of `predict()`). The data frame requires data from _all_ model terms, including random effect terms. This again requires to choose certain levels or values also for each random effect term, or to set those terms to zero or `NA` (for population-level). Since there is no general rule, which level(s) of random effect terms to choose in order to represent the random effects structure in the data, using the population-level seems the most clear and consistent approach.

To get predicted values for a specific level of the random effect term, simply define this level in the `condition`-argument.

```{r}
ggpredict(m, "Days", type = "random", condition = c(Subject = 330))
```

Finally, it is possible to obtain predicted values by simulating from the model, where predictions are based on `simulate()`.
 
```{r}
ggpredict(m, "Days", type = "simulate")
```

## When are predictions affected by `type = "fixed"` and `type = "random"`?

When `type = "fixed"`, `ggpredict()` calls `predict(..., re.form = NA)`. For `type = "random"`, `predict(..., re.form = NULL)` is called. This can lead to different predictions, depending on whether `REML = TRUE` or `REML = FALSE` was used for model fitting. When `REML = FALSE`, `re.form = NA` and `re.form = NULL` are identical, and thus predictions are not affected by `type = "fixed"` and `type = "random"`. However, when `REML = TRUE`, `re.form = NA` and `re.form = NULL` can return different predictions, also depending on whether factors are included in the model or not. The following example shows a case where predictions are affected by `type = "fixed"` and `type = "random"`.

```{r}
library(glmmTMB)
set.seed(123)
sleepstudy$x <- as.factor(sample(1:3, nrow(sleepstudy), replace = TRUE))
# REML is FALSE
m1 <- glmmTMB(Reaction ~ Days + x + (1 + Days | Subject), data = sleepstudy, REML = FALSE)
# REML is TRUE
m2 <- glmmTMB(Reaction ~ Days + x + (1 + Days | Subject), data = sleepstudy, REML = TRUE)

# predictions when REML is FALSE - no difference between type = "fixed"
# and type = "random" in predictions, only for intervals
ggpredict(m1, "Days [1:3]")
ggpredict(m1, "Days [1:3]", type = "random")

# predictions when REML is TRUE - we now see a difference both
# for intervals *and* predictions
ggpredict(m2, "Days [1:3]")
ggpredict(m2, "Days[1:3]", type = "random")
```

## To summarize...

- `type = "fixed"`: predictions are on the population-level, and do not account for the random effect variances. Intervals are _confidence intervals_ for the predicted values. `re.form = NA` when calling `predict()`.

- `type = "random"`: predictions are on the population-level, but conditioned on the random effects (individual level predictions), which means that `re.form = NULL` when calling `predict()`. Intervals are _prediction intervals_.

- `type = "random", interval = "confidence"`: predictions are on the population-level, conditioned on the random effects (which means that `re.form = NULL` when calling `predict()`), however, intervals are _confidence intervals_.

## Marginal effects for zero-inflated mixed models

For zero-inflated mixed effects models, typically fitted with the **glmmTMB** or **GLMMadaptive** packages, predicted values can be conditioned on

  * the fixed effects of the conditional model only (`type = "fixed"`)
  * the fixed effects and zero-inflation component (`type = "zero_inflated"`)
  * the fixed effects of the conditional model only (population-level), taking the random-effect variances into account (`type = "random"`)
  * the fixed effects and zero-inflation component (population-level), taking the random-effect variances into account (`type = "zi_random"`)
  * all model parameters (`type = "simulate"`)

```{r}
library(glmmTMB)
data(Salamanders)
m <- glmmTMB(
  count ~ spp + mined + (1 | site),
  ziformula = ~ spp + mined,
  family = truncated_poisson,
  data = Salamanders
)
```

Similar to mixed models without zero-inflation component, `type = "fixed"` and `type = "random"` for **glmmTMB**-models (with zero-inflation) both return predictions on the population-level, where the latter option accounts for the uncertainty of the random effects. In short, `predict(..., type = "link")` is called (however, predicted values are back-transformed to the response scale).

```{r}
ggpredict(m, "spp")

ggpredict(m, "spp", type = "random")
```

For `type = "zero_inflated"`, the predicted response value is the expected value `mu*(1-p)` _without conditioning_ on random effects. Since the zero inflation and the conditional model are working in "opposite directions", a higher expected value for the zero inflation means a lower response, but a higher value for the conditional model means a higher response. While it is possible to calculate predicted values with `predict(..., type = "response")`, standard errors and confidence intervals can not be derived directly from the `predict()`-function. Thus, confidence intervals for `type = "zero_inflated"` are based on quantiles of simulated draws from a multivariate normal distribution (see also _Brooks et al. 2017, pp.391-392_ for details).


```{r}
ggpredict(m, "spp", type = "zero_inflated")
```

For `type = "zi_random"`, the predicted response value is the expected value `mu*(1-p)`, accounting for the random-effect variances. Intervals are calculated in the same way as for `type = "zero_inflated"`, except that the mean random effect variance is considered and thus _prediction intervals_ rather than confidence intervals are returned.

```{r}
ggpredict(m, "spp", type = "zi_random")
```

Finally, it is possible to obtain predicted values by simulating from the model, where predictions are based on `simulate()` (see _Brooks et al. 2017, pp.392-393_ for details). To achieve this, use `type = "simulate"`.
 
```{r}
ggpredict(m, "spp", type = "simulate")
```

## Marginal effects for each level of random effects

Marginal effects can also be calculated for each group level in mixed models. Simply add the name of the related random effects term to the `terms`-argument, and set `type = "random"`.

In the following example, we fit a linear mixed model and first simply plot the marginal effects, _not_ conditioned on random-effect variances.

```{r}
library(sjlabelled)
data(efc)
efc$e15relat <- as_label(efc$e15relat)
m <- lmer(neg_c_7 ~ c12hour + c160age + c161sex + (1 | e15relat), data = efc)
me <- ggpredict(m, terms = "c12hour")
plot(me)
```

Changing the type to `type = "random"` still returns population-level predictions by default. Recall that the major difference between `type = "fixed"` and `type = "random"` is the uncertainty in the variance parameters. This leads to larger confidence intervals (i.e. prediction intervals) for marginal effects with `type = "random"`.

```{r}
me <- ggpredict(m, terms = "c12hour", type = "random")
plot(me)
```

To compute marginal effects for each grouping level, add the related random term to the `terms`-argument. In this case, prediction intervals are calculated and marginal effects are conditioned on each group level of the random effects.

```{r}
me <- ggpredict(m, terms = c("c12hour", "e15relat"), type = "random")
plot(me, show_ci = FALSE)
```

> When predictions for group levels (random effects) are of interest, it can be useful to calculate confidence instead of prediction intervals (i.e. to ignore the uncertainty in the variance parameters and return regular confidence intervals), in particular, when `hypothesis_test()` is used for contrasts or comparisons of group levels. In that case, set `interval = "confidence"` (see examples below).

Marginal effects, conditioned on random effects, can also be calculated for specific levels only. Add the related values into brackets after the variable name in the `terms`-argument.

```{r}
me <- ggpredict(m, terms = c("c12hour", "e15relat [child,sibling]"), type = "random")
plot(me, show_ci = FALSE)
```

...and including prediction intervals...

```{r}
plot(me)
```

The most complex plot in this scenario would be a term (`c12hour`) at certain values of two other terms (`c161sex`, `c160age`) for specific levels of random effects (`e15relat`), so we have four variables in the `terms`-argument.

```{r fig.height=6}
me <- ggpredict(
  m,
  terms = c("c12hour", "c161sex", "c160age", "e15relat [child,sibling]"), 
  type = "random"
)
plot(me)
```

If the group factor has too many levels, you can also take a random sample of all possible levels and plot the marginal effects for this subsample of group levels. To do this, use `term = "<groupfactor> [sample=n]"`.

```{r}
set.seed(123)
m <- lmer(Reaction ~ Days + (1 + Days | Subject), data = sleepstudy)
me <- ggpredict(m, terms = c("Days", "Subject [sample=7]"), type = "random")
plot(me)
```

You can also add the observed data points for each group using `show_data = TRUE`.

```{r}
plot(me, show_data = TRUE, show_ci = FALSE)
```

### Prediction and usual confidence intervals

If random effect levels are of interest, setting `type = "random"` is obligatory. However, sometimes it can be useful to have usual confidence intervals instead of prediction intervals, e.g. for [pairwise comparisons](introduction_comparisons_1) of random effects. Confidence instead of prediction intervals can be calculated by explicitly setting `interval = "confidence"`, in which case the random effects variances are ignored.

```{r}
me <- ggpredict(
  m,
  terms = c("Days", "Subject [sample=7]"),
  type = "random",
  interval = "confidence"
)
plot(me)
```

# References

Brooks ME, Kristensen K, Benthem KJ van, Magnusson A, Berg CW, Nielsen A, et al. glmmTMB Balances Speed and Flexibility Among Packages for Zero-inflated Generalized Linear Mixed Modeling. The R Journal. 2017;9: 378–400.

Johnson PC, O'Hara RB. 2014. Extension of Nakagawa & Schielzeth's R2GLMM to random slopes models. Methods Ecol Evol, 5: 944-946. (doi: 10.1111/2041-210X.12225)
