---
title: "Significance Testing of Differences Between Predictions III: Contrasts and Comparisons for Generalized Linear Models"
author: "Daniel Lüdecke"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Significance Testing of Differences Between Predictions III: Contrasts and Comparisons for Generalized Linear Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r set-options, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "png",
  fig.width = 7,
  fig.height = 3.5,
  message = FALSE, warning = FALSE)
options(width = 800)
arrow_color <- "#FF00cc"

if (!requireNamespace("ggplot2", quietly = TRUE) || !requireNamespace("marginaleffects", quietly = TRUE)) {
  knitr::opts_chunk$set(eval = FALSE)
}
```

This vignette is the third in a 3-part series:

1. [**Significance Testing of Differences Between Predictions I: Contrasts and Pairwise Comparisons**](https://strengejacke.github.io/ggeffects/articles/introduction_comparisons_1.html)

2. [**Significance Testing of Differences Between Predictions II: Comparisons of Slopes, Floodlight and Spotlight Analysis (Johnson-Neyman Intervals)**](https://strengejacke.github.io/ggeffects/articles/introduction_comparisons_2.html)

3. **Significance Testing of Differences Between Predictions III: Contrasts and Comparisons for Generalized Linear Models**


# Contrasts and comparisons for GLM - logistic regression example

Lastly, we show an example for non-Gaussian models. For GLM models with (non-Gaussian) link-functions, `predict_response()` always returns predcted values on the *response* scale. For example, predicted values for logistic regression models are shown as *probabilities*.

Let's look at a simple example

```{r}
library(ggeffects)
set.seed(1234)
dat <- data.frame(
  outcome = rbinom(n = 100, size = 1, prob = 0.35),
  x1 = as.factor(sample(1:3, size = 100, TRUE, prob = c(0.5, 0.2, 0.3))),
  x2 = rnorm(n = 100, mean = 10, sd = 7)
)

m <- glm(outcome ~ x1 + x2, data = dat, family = binomial())
predict_response(m, "x1")
```

## Contrasts and comparisons for categorical focal terms

Contrasts or comparisons - like predictions (see above) - are by default on the *response* scale, i.e. they're represented as difference between probabilities (in percentage points).

```{r message=TRUE}
p <- predict_response(m, "x1")
hypothesis_test(p)
```

```{r echo=FALSE}
ht8 <- hypothesis_test(p)
```

The difference between the predicted probability of `x1 = 1` (`r sprintf("%.1f%%", 100 * p$predicted[1])`) and `x1 = 2` (`r sprintf("%.1f%%", 100 * p$predicted[2])`) is roughly `r sprintf("%.1f%%", 100 * ht8$Contrast[1])` points. This difference is not statistically significant (p = `r round(ht8$p.value[1], 3)`).

The `scale` argument in `hypothesis_test()` can be used to return contrasts or comparisons on a differen scale. For example, to transform contrasts to _odds ratios_, we can use `scale = "exp"`.

```{r message=TRUE}
hypothesis_test(p, scale = "exp")
```

Contrasts or comparisons can also be represented on the link-scale, in this case as _log-odds_. To do so, use `scale = "link"`.

```{r message=TRUE}
hypothesis_test(p, scale = "link")
```

## Contrasts and comparisons for numerical focal terms

For numeric focal variables, where the slopes (linear trends) are estimated, transformed scales (like `scale = "exp"`) are not supported. However, `scale = "link"` can be used to return untransformed contrasts or comparisons on the link-scale.

```{r message=TRUE}
hypothesis_test(m, "x2", scale = "link")
```

Be aware whether and which back-transformation to use, as it affects the resulting p-values. A detailed overview of transformations can be found [in this vignette](https://vincentarelbundock.github.io/marginaleffects/articles/comparisons.html).
