---
title: "Different backends for pairwise comparisons"
author: "Daniel Lüdecke"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Different backends for pairwise comparisons}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r set-options, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "png",
  fig.width = 7,
  fig.height = 3.5,
  message = FALSE, warning = FALSE)
options(width = 800)
arrow_color <- "#FF00cc"

if (!requireNamespace("ggplot2", quietly = TRUE) ||
    !requireNamespace("marginaleffects", quietly = TRUE) ||
    !requireNamespace("parameters", quietly = TRUE)) {
  knitr::opts_chunk$set(eval = FALSE)
}
```

This vignette is roughly a duplication of the first vignette about [*Contrasts and Pairwise Comparisons*](https://strengejacke.github.io/ggeffects/articles/introduction_comparisons_1.html), but demonstrating the different backends for the calculation of pairwise comparisons. The default backend is the [*marginaleffects*](https://vincentarelbundock.github.io/marginaleffects/) package. If desired, `engine = "emmeans"` can be used to switch to the [*emmeans*](https://cran.r-project.org/package=emmeans) package. `engine = "ggeffects"` is currently experimental and work-in-progress. It is less feature-rich than the other backends, but it also works for predictions of random effects.

## Within `episode`, do levels differ?

```{r}
library(ggeffects)
library(ggplot2)

set.seed(123)
n <- 200
d <- data.frame(
  outcome = rnorm(n),
  grp = as.factor(sample(c("treatment", "control"), n, TRUE)),
  episode = as.factor(sample(1:3, n, TRUE)),
  sex = as.factor(sample(c("female", "male"), n, TRUE, prob = c(0.4, 0.6)))
)
model1 <- lm(outcome ~ grp + episode + grp, data = d)
```

### Predictions

```{r}
my_pred <- predict_response(model1, "episode", margin = "marginalmeans")
my_pred
```

### Pairwise comparisons

```{r echo=FALSE}
p <- plot(my_pred)
line_data <- as.data.frame(my_pred, terms_to_colnames = FALSE)[1:2, ]
p + geom_segment(
  data = line_data,
  aes(
    x = as.numeric(x[1]) + 0.06, xend = as.numeric(x[2]) - 0.06,
    y = predicted[1], yend = predicted[2], group = NULL, color = NULL
  ),
  color = arrow_color,
  arrow = arrow(length = unit(0.1, "inches"), ends = "both", angle = 40)
) +
ggtitle("Within \"episode\", do levels 1 and 2 differ?")
```

```{r}
# comparisons based on estimated marginal means, using "marginaleffects" package
test_predictions(model1, "episode", margin = "marginalmeans")

# comparisons using "emmeans" package
test_predictions(model1, "episode", engine = "emmeans")

# comparisons using "ggeffects" backend. This engine requires the
# ggeffects-object as input
test_predictions(my_pred, engine = "ggeffects")
```


## Does same level of episode differ between groups?

```{r}
model2 <- lm(outcome ~ grp * episode + grp, data = d)
```

### Predictions

```{r}
my_pred <- predict_response(model2, c("episode", "grp"), margin = "marginalmeans")
my_pred
```

### Pairwise comparisons

```{r echo=FALSE}
p <- plot(my_pred)
line_data <- as.data.frame(my_pred, terms_to_colnames = FALSE)[3:4, 1:2]
line_data$group_col <- "control"
p + geom_segment(
  data = line_data,
  aes(
    x = as.numeric(x[1]) - 0.06, xend = as.numeric(x[2]) + 0.06,
    y = predicted[1], yend = predicted[2], group = NULL, color = NULL
  ),
  color = arrow_color,
  arrow = arrow(length = unit(0.1, "inches"), ends = "both", angle = 40)
) +
ggtitle("Within level 2 of \"episode\", do treatment and control group differ?")
```

```{r}
# we want "episode = 2-2" and "grp = control-treatment"

# comparisons based on estimated marginal means, using "marginaleffects" package
test_predictions(model2, c("episode [2]", "grp"), margin = "marginalmeans")

# comparisons based using "emmeans" package
test_predictions(model2, c("episode [2]", "grp"), engine = "emmeans")

# comparisons using "ggeffects" backend
my_pred <- predict_response(model2, c("episode [2]", "grp"), margin = "marginalmeans")
test_predictions(my_pred, engine = "ggeffects")
```

## Does difference between two levels of episode in the control group differ from difference of same two levels in the treatment group?

The `test` argument also allows us to compare difference-in-differences. When `engine = "emmeans"` or `"ggeffects"`, we need to set `test = "interaction"` to get interaction contrasts, i.e. differences-in-differences.

```{r echo=FALSE}
my_pred <- predict_response(model2, c("grp", "episode"))
p <- plot(my_pred)
line_data <- as.data.frame(my_pred, terms_to_colnames = FALSE)[, 1:2, ]
line_data$group_col <- "1"
p + geom_segment(
  data = line_data,
  aes(
    x = as.numeric(x[1]) - 0.05, xend = as.numeric(x[1]) - 0.05,
    y = predicted[1], yend = predicted[2], group = NULL, color = NULL
  ),
  color = "orange",
  arrow = arrow(length = unit(0.1, "inches"), ends = "both", angle = 40, type = "closed")
) + geom_segment(
  data = line_data,
  aes(
    x = as.numeric(x[4]) - 0.05, xend = as.numeric(x[4]) - 0.05,
    y = predicted[4], yend = predicted[5], group = NULL, color = NULL
  ),
  color = "orange",
  arrow = arrow(length = unit(0.1, "inches"), ends = "both", angle = 40, type = "closed")
) + geom_segment(
  data = line_data,
  aes(
    x = as.numeric(x[1]) - 0.05, xend = as.numeric(x[4]) - 0.05,
    y = (predicted[1] + predicted[2]) / 2,
    yend = (predicted[4] + predicted[5]) / 2, group = NULL, color = NULL
  ),
  color = arrow_color,
  arrow = arrow(length = unit(0.1, "inches"), ends = "both", angle = 40)
) +
ggtitle("Differnce-in-differences")
```

```{r}
# specifying the difference-in-difference when using "marginaleffects"
test_predictions(model2, c("episode", "grp"), test = "(b1 - b3) = (b2 
- b4)", margin = "marginalmeans")

# using "emmeans", we need to set "test = "interaction". This displays *all*
# possible differences-in-differences. The first row in this output is identical
# to the above result from "marginaleffects"
test_predictions(model2, c("episode", "grp"), test = "interaction", engine = "emmeans")

# using "ggeffects", we also need to set "test = "interaction"
my_pred <- predict_response(model2, c("episode", "grp"), margin = "marginalmeans")
test_predictions(my_pred, engine = "ggeffects")
```
