---
title: "Introduction: Adjusted Predictions and Marginal Effects for Random Effects Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction: Adjusted Predictions and Marginal Effects for Random Effects Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r set-options, echo=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "png",
  fig.width = 7,
  fig.height = 4,
  message = FALSE,
  warning = FALSE
)
options(width = 800, ggeffects_warning_bias_correction = FALSE)

pkgs <- c(
  "ggplot2",
  "lme4",
  "glmmTMB",
  "patchwork",
  "sjlabelled",
  "htmltools"
)

if (!all(vapply(pkgs, requireNamespace, quietly = TRUE, FUN.VALUE = logical(1L)))) {
  knitr::opts_chunk$set(eval = FALSE)
}
```

```{r echo=FALSE, message=FALSE}
library(htmltools)
callout_tip <- function(header = NULL, ...) {
  div(
    class = "callout-tip",
    tags$h1(
      tags$img(src = "../man/figures/summary.png", width = "20", height = "17", style = "vertical-align:middle"), # nolint
      header
    ),
    ...
  )
}
includeCSS("../man/figures/callout.css")
```

This vignette shows how to calculate adjusted predictions for mixed models. However, for mixed models, since random effects are involved, we can calculate _conditional predictions_ and _marginal predictions_. We also have to distinguish between _population-level_ and _unit-level_ predictions.

But one thing at a time...

```{r echo=FALSE}
callout_tip(
  "Summary of most important points:",
  tags$ul(
    tags$li("Predictions can be made on the population-level or for each level of the grouping variable (unit-level). If unit-level predictions are requested, you need to set ", tags$code("type=\"random\""), " and specify the grouping variable(s) in the ", tags$code("terms"), " argument."), # nolint
    tags$li("Population-level predictions can be either conditional (predictions for a \"typical\" group) or marginal (average predictions across all groups). Set ", tags$code("margin=\"empirical\""), " for marginal predictions. You'll notice differences in predictions especially for unequal group sizes at the random effects level."), # nolint
    tags$li("Prediction intervals, i.e. when ", tags$code("interval=\"predictions\""), " also account for the uncertainty in the random effects.") # nolint
  )
)
```

## Population-level predictions for mixed effects models

Mixed models are used to account for the dependency of observations within groups, e.g. repeated measurements within subjects, or students within schools. The dependency is modeled by random effects, i.e. mixed model at least have one grouping variable (or factor) as higher level unit.

At the lowest level, you have your _fixed effects_, i.e. your "independent variables" or "predictors".

Adjusted predictions can now be calculated for specified values or levels of the focal terms, however, either for the full sample (population-level) or for each level of the grouping variable (unit-level). The latter is particularly useful when the grouping variable is of interest, e.g. when you want to compare the effect of a predictor between different groups.

To get population-level predictions, we set `type = "fixed"` or `type = "zero_inflation"` (for models with zero-inflation component).

### Conditional and marginal effects and predictions

We start with the population-level predictions. Here you can either calculate the _conditional_ or the _marginal_ effect (see in detail also _Heiss 2022_). The conditional effect is the effect of a predictor in an average or typical group, while the marginal effect is the average effect of a predictor across all groups. E.g. let's say we have `countries` as grouping variable and `gdp` (gross domestic product per capita) as predictor, then the conditional and marginal effect would be:

- conditional effect: effect of `gdp` in an _average_ or _typical_ country. To get conditional predictions, we use `predict_response()` or `predict_response(margin = "mean_mode")`.

- marginal effect: average effect of `gdp` _across all_ countries. To get marginal (or average) predictions, we use `predict_response(margin = "empirical")`.

While the term "effect" referes to the strength of the relationship between a predictor and the response, "predictions" refer to the actual predicted values of the response. Thus, in the following, we will talk about conditional and marginal (or average) _predictions_.

In a balanced data set, where all groups have the same number of observations, the conditional and marginal predictions are often similar (maybe slightly different, depending on the non-focal predictors). However, in unbalanced data, the conditional and marginal predicted values can largely differ.

```{r}
library(ggeffects)
library(lme4)
data(sleepstudy)

# balanced data set
m <- lmer(Reaction ~ Days + (1 + Days | Subject), data = sleepstudy)

# conditional predictions
predict_response(m, "Days [1,5,9]")

# average marginal predictions
predict_response(m, "Days [1,5,9]", margin = "empirical")

# create imbalanced data set
set.seed(123)
strapped <- sleepstudy[sample.int(nrow(sleepstudy), nrow(sleepstudy), replace = TRUE), ]
m <- lmer(Reaction ~ Days + (1 + Days | Subject), data = strapped)

# conditional predictions
predict_response(m, "Days [1,5,9]")

# average marginal predictions
predict_response(m, "Days [1,5,9]", margin = "empirical")
```

### Population-level predictions and the `REML` argument

The conditional predictions returned by `predict_response()` for the default marginalization (i.e. when `margin` is set to `"mean_reference"` or `"mean_mode"`) may differ...

- depending on the setting of the `REML` argument during model fitting;
- and depending on whether factors are included in the model or not.

```{r}
library(glmmTMB)
set.seed(123)
sleepstudy$x <- as.factor(sample(1:3, nrow(sleepstudy), replace = TRUE))
# REML is FALSE
m1 <- glmmTMB(Reaction ~ Days + x + (1 + Days | Subject), data = sleepstudy, REML = FALSE)
# REML is TRUE
m2 <- glmmTMB(Reaction ~ Days + x + (1 + Days | Subject), data = sleepstudy, REML = TRUE)

# predictions when REML is FALSE
predict_response(m1, "Days [1:3]")

# predictions when REML is TRUE
predict_response(m2, "Days [1:3]")
```

### Summary of population-level predictions

For *conditional* predictions (i.e. the default marginalization method in `predict_response()`), we can summarize how they work for mixed models like this:

- `type = "fixed"`: Predictions are on the population-level (`re.form = NA` when calling `predict()`). This is the default and recommended option for population-level prediction in mixed models.

- `type = "random"`: if *no* random effect terms (grouping variables) are included in the `terms` argument, predictions are still on the population-level, but conditioned on the random effects (`re.form = NULL` when calling `predict()`). This may return different predictions than `type = "fixed"`, in particular for non-Gaussian models.

*Marginal* predictions can be obtained by setting `margin = "empirical"`.

## Population-level predictions for zero-inflated mixed models

For zero-inflated mixed effects models, typically fitted with the **glmmTMB** or **GLMMadaptive** packages, `predict_response()` can return predicted values of the response, conditioned on following prediction-types:

  - Conditional effects / predictions:

    * population-level predictions, conditioned on the fixed effects of the conditional (or "count") model only (`type = "fixed"`)
    * population-level predictions, conditioned on the fixed effects *and* zero-inflation component (`type = "zero_inflated"`), returning the expected (predicted) values of the response
    * the zero-inflation probabilities (`type = "zi_prob"`)

  - Marginal effects / predictions:

    * `type = "simulate"` or `margin = "empirical"` can be used to obtain marginal predictions again.

 For `predict_response(margin = "empirical")`, `type = "simulate"` is not available. However, valid values for `type` can also be those based on the model's `predict()` method. For models of class `glmmTMB`, these are `"response"`, `"link"`, `"conditional"`, `"zprob"`, `"zlink"`, or `"disp"`.

### Conditional predictions: Adjusted predictions for the conditional model

For now, we show examples for conditional predictions, which is the default marginalization method in `predict_response()`.

```{r}
library(glmmTMB)
data(Salamanders)
m <- glmmTMB(
  count ~ spp + mined + (1 | site),
  ziformula = ~ spp + mined,
  family = poisson(),
  data = Salamanders
)
```

Similar to mixed models without zero-inflation component, `type = "fixed"` returns predictions on the population-level, but for the conditional model only.

```{r}
predict_response(m, "spp")
```

### Conditional predictions: Adjusted predictions for the full model

For `type = "zero_inflated"`, the predicted response value is the expected value `mu*(1-p)`. Since the zero inflation and the conditional model are working in "opposite directions", a higher expected value for the zero inflation means a lower response, but a higher value for the conditional ("count") model means a higher response. While it is possible to calculate predicted values with `predict(..., type = "response")`, standard errors and confidence intervals can not be derived directly from the `predict()`-function. Thus, confidence intervals for `type = "zero_inflated"` are based on quantiles of simulated draws from a multivariate normal distribution (see also _Brooks et al. 2017, pp.391-392_ for details).

```{r}
predict_response(m, "spp", type = "zero_inflated")
```

### Marginal predictions: Simulated outcome (full model)

In the above examples, we get the conditional, not the marginal predictions. E.g., predictions are conditioned on `mined` when it is set to `"yes"`. However, it is possible to obtain predicted values by simulating from the model, where predictions are based on `simulate()` (see _Brooks et al. 2017, pp.392-393_ for details). This will return expected values of the response (*marginal* predictions), averaged across all random effects groups and non-focal terms. To achieve this, use `type = "simulate"`. Note that intervals of this type will be wider than those of `type = "zero_inflated"`, because they account for the full uncertainty of model parameters. It is even possible to return prediction intervals using `interval = "prediction"` when `type = "simulate"`.

```{r}
predict_response(m, "spp", type = "simulate")
```

### Marginal predictions: Average predictions for the full model

In a similar fashion, you can obtain average marginal predictions for zero-inflated mixed models with `margin = "empirical"`. The returned values are most comparable to `predict_response(type = "simulate")`, because `margin = "empirical"` also returns expected values of the response, averaged across all random effects groups and all non-focal terms. The next example shows the average marginal predicted values of `spp` on the response across all `site`s, taking the zero-inflation component into account (i.e. `type = "zero_inflated"`).

```{r}
predict_response(m, "spp", type = "zero_inflated", margin = "empirical")
```

## Bias-correction for non-Gaussian models

For non-Gaussian models, predicted values are back-transformed to the response scale. However, back-transforming the population-level predictions (in _mixed_ models, when `type = "fixed"`) ignores the effect of the variation around the population mean, hence, the result on the original data scale is biased due to _Jensen's inequality_. In this case, it can be appropriate to apply a bias-correction. This is done by setting `bias_correction = TRUE`. By default, [`insight::get_variance_residual()`](https://easystats.github.io/insight/reference/get_variance_residual.html) is used to extract the residual variance, which is used to calculate the amount of correction. Optionally, you can provide your own estimates of uncertainty, e.g. based on [`insight::get_variance_random()`](https://easystats.github.io/insight/reference/get_variance_random.html), using the `sigma` argument.

Bias-correction comes closer to the expected values of the response, averaged across all random effects groups and all non-focal terms. The following example shows the bias-corrected predicted values of `spp`, which are closer to the simulated values than the non-bias-corrected ones.

```{r, eval=TRUE, echo=FALSE}
options(ggeffects_warning_bias_correction = TRUE)
```

```{r}
# no bias-correction
predict_response(m, "spp")

# bias-correction
predict_response(m, "spp", bias_correction = TRUE)

# closer to "simulate" for some of the predicted values
predict_response(m, "spp", type = "simulate")

# bias-correction, using user-defined sigma-value
predict_response(m, "spp", bias_correction = TRUE, sigma = insight::get_sigma(m))
```

## Unit-level predictions (predictions for each level of random effects)

Adjusted predictions can also be calculated for each group level (unit-level) in mixed models. Simply add the name of the related random effects term to the `terms`-argument, and set `type = "random"`. For `predict_response(margin = "empirical")`, you don't need to set `type = "random"`.

In the following example, we fit a linear mixed model and first simply plot the adjusted predictions, _not_ conditioned on random-effect variances.

```{r}
library(sjlabelled)
data(efc)
efc$e15relat <- as_label(efc$e15relat)
m <- lmer(neg_c_7 ~ c12hour + c160age + c161sex + (1 | e15relat), data = efc)
me <- predict_response(m, terms = "c12hour")
plot(me)
```

To compute adjusted predictions for each grouping level, add the related random term to the `terms`-argument. In this case, predictions are conditioned on each unit-level of the random effects.

```{r}
me <- predict_response(m, terms = c("c12hour", "e15relat"), type = "random")
plot(me, show_ci = FALSE)
```

Since average marginal predictions already consider random effects by averaging over the groups, the `type`-argument is not needed when `margin = "empirical"` is set.

```{r}
me <- predict_response(m, terms = c("c12hour", "e15relat"), margin = "empirical")
plot(me, show_ci = FALSE)
```

Adjusted predictions, conditioned on random effects, can also be calculated for specific unit-levels only. Add the related values into brackets after the variable name in the `terms`-argument.

```{r}
me <- predict_response(m, terms = c("c12hour", "e15relat [child,sibling]"), type = "random")
plot(me, show_ci = FALSE)
```

The most complex plot in this scenario would be a term (`c12hour`) at certain values of two other terms (`c161sex`, `c160age`) for specific unit-levels of random effects (`e15relat`), so we have four variables in the `terms`-argument.

```{r fig.height=7}
me <- predict_response(
  m,
  terms = c("c12hour", "c161sex", "c160age", "e15relat [child,sibling]"),
  type = "random"
)
plot(me, n_rows = 2)
```

If the group factor has too many levels, you can also take a random sample of all possible levels and plot the adjusted predictions for this subsample of unit-levels. To do this, use `term = "<groupfactor> [sample=n]"`.

```{r}
set.seed(123)
m <- lmer(Reaction ~ Days + (1 + Days | Subject), data = sleepstudy)
me <- predict_response(m, terms = c("Days", "Subject [sample=7]"), type = "random")
plot(me)
```

You can also add the observed data points for each group using `show_data = TRUE`.

```{r}
plot(me, show_data = TRUE, show_ci = FALSE)
```

## Population-level predictions for `gam` and `glmer` models

The output of `predict_response()` indicates that the grouping variable of the random effects is set to "population level" (adjustment), e.g. in case of *lme4*, following is printed:

> Adjusted for:
> * Subject = 0 (population-level)

A comparable model fitted with `mgcv::gam()` would print a different message:

> Adjusted for:
> * Subject = 308

The reason is because the correctly printed information about adjustment for random effects is based on `insight::find_random()`, which returns `NULL` for `gam`s with random effects defined via `s(..., bs = "re")`. However, predictions are still correct, when population-level predictions are requested. Here's an example:

```{r message = FALSE}
data("sleepstudy", package = "lme4")
# mixed model with lme4
m_lmer <- lme4::lmer(Reaction ~ poly(Days, 2) + (1 | Subject),
  data = sleepstudy
)
# equivalent model, random effects are defined via s(..., bs = "re")
m_gam <- mgcv::gam(Reaction ~ poly(Days, 2) + s(Subject, bs = "re"),
  family = gaussian(), data = sleepstudy, method = "ML"
)

# predictions are identical
predict_response(m_gam, terms = "Days", exclude = "s(Subject)", newdata.guaranteed = TRUE)

predict_response(m_lmer, terms = "Days")
```

# References

Brooks ME, Kristensen K, Benthem KJ van, Magnusson A, Berg CW, Nielsen A, et al. glmmTMB Balances Speed and Flexibility Among Packages for Zero-inflated Generalized Linear Mixed Modeling. The R Journal. 2017;9: 378–400.

Heiss, A. (2022, November 29). Marginal and conditional effects for GLMMs with {marginaleffects}. Andrew Heiss’s Blog. (doi: 10.59350/xwnfm-x1827)

Johnson PC. 2014. Extension of Nakagawa & Schielzeth's R2GLMM to random slopes models. Methods Ecol Evol, 5: 944-946. (doi: 10.1111/2041-210X.12225)
