---
title:"Significance Testing of Differences Between Predictions: Slopes, Contrasts and Pairwise Comparisons"
author: "Daniel Lüdecke"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Significance Testing of Differences Between Predictions: Slopes, Contrasts and Pairwise Comparisons}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r set-options, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "png",
  fig.width = 7,
  fig.height = 3.5,
  message = FALSE, warning = FALSE)
options(width = 800)

if (!requireNamespace("ggplot2", quietly = TRUE) ||
    !requireNamespace("gridExtra", quietly = TRUE) ||
    !requireNamespace("see", quietly = TRUE) ||
    !requireNamespace("patchwork", quietly = TRUE) ||
    !requireNamespace("sjmisc", quietly = TRUE)) {
  knitr::opts_chunk$set(eval = FALSE)
}
```

# Aim of the ggeffects-package

```{r}
library(ggeffects)
library(ggplot2)

data(efc)
model <- lm(barthtot ~ c12hour + neg_c_7 + c161sex + c172code, data = efc)
mydf <- ggpredict(model, terms = "c12hour")

theme_set(theme_bw())
ggplot(mydf, aes(x, predicted)) +
  geom_line()
```

As you can see, `ggpredict()` (and `ggeffect()` or `ggemmeans()`) has a nice `print()`-method, which takes care of printing not too many rows (but always an equally spaced range of values, including minimum and maximum value of the term in question) and giving some extra information. This is especially useful when predicted values are shown depending on the levels of other terms (see below).

The output shows the predicted values for the response at each value from the term _c12hour_. The data is already in shape for ggplot:

```{r}
# automatic detection of numeric focal variable
# calculating p-value for linear trend 
ggcomparisons(model, "c12hour")
```

# Marginal effects at the mean by other predictors' levels

The `terms`-argument accepts up to four model terms, where the second to fourth terms indicate grouping levels. This allows predictions for the term in question at different levels for other model terms:

```{r}
ggpredict(fit, terms = c("c12hour", "c172code"))
```

Creating a ggplot is pretty straightforward: the `colour`-aesthetics is mapped with the `group`-column:

```{r}
mydf <- ggpredict(fit, terms = c("c12hour", "c172code"))
ggplot(mydf, aes(x, predicted, colour = group)) + geom_line()
```

A second grouping structure can be defined, which will create another column named `facet`, which - as the name implies - might be used to create a facted plot:

```{r}
mydf <- ggpredict(fit, terms = c("c12hour", "c172code", "c161sex"))
mydf
ggplot(mydf, aes(x, predicted, colour = group)) + 
  geom_line() + 
  facet_wrap(~facet)
```

Finally, a third differentation can be defined, creating another column named `panel`. In such cases, you may create multiple plots (for each value in `panel`). **ggeffects** takes care of this when you use `plot()` and automatically creates an integrated plot with all panels in one figure.

```{r fig.height = 8}
mydf <- ggpredict(fit, terms = c("c12hour", "c172code", "c161sex", "neg_c_7"))
plot(mydf)
```

# Marginal effects for each model term

If the `term` argument is either missing or `NULL`, marginal effects for each model term are calculated. The result is returned as a list, which can be plotted manually (or using the `plot()` function).

```{r}
mydf <- ggpredict(fit)
mydf
```


# Two-Way, Three-Way- and Four-Way-Interactions

To plot the marginal effects of interaction terms, simply specify these terms in the `terms`-argument.

```{r}
library(sjmisc)
data(efc)

# make categorical
efc$c161sex <- to_factor(efc$c161sex)

# fit model with interaction
fit <- lm(neg_c_7 ~ c12hour + barthtot * c161sex, data = efc)

# select only levels 30, 50 and 70 from continuous variable Barthel-Index
mydf <- ggpredict(fit, terms = c("barthtot [30,50,70]", "c161sex"))
ggplot(mydf, aes(x, predicted, colour = group)) + geom_line()
```

Since the `terms`-argument accepts up to four model terms, you can also compute marginal effects for a 3-way-interaction or 4-way-interaction. To plot the marginal effects of three interaction terms, just like before, specify all three terms in the `terms`-argument.

```{r}
# fit model with 3-way-interaction
fit <- lm(neg_c_7 ~ c12hour * barthtot * c161sex, data = efc)

# select only levels 30, 50 and 70 from continuous variable Barthel-Index
mydf <- ggpredict(fit, terms = c("c12hour", "barthtot [30,50,70]", "c161sex"))

ggplot(mydf, aes(x, predicted, colour = group)) + 
  geom_line() +
  facet_wrap(~facet)
```

4-way-interactions are rather confusing to print and plot. When plotting, multiple plots (for each level of the fourth interaction term) are plotted for the remaining three interaction terms. This can easily be done using the [`plot()`-method](https://strengejacke.github.io/ggeffects/articles/introduction_plotmethod.html).

```{r fig.height = 8}
# fit model with 4-way-interaction
fit <- lm(neg_c_7 ~ c12hour * barthtot * c161sex * c172code, data = efc)

# marginal effects for all 4 interaction terms
pr <- ggpredict(fit, c("c12hour", "barthtot", "c161sex", "c172code"))

# use plot() method, easier than own ggplot-code from scratch
plot(pr)
```


# Polynomial terms and splines

`ggpredict()` also works for models with polynomial terms or splines. Following code reproduces the plot from `?splines::bs`:


```{r}
library(splines)
data(women)

fm1 <- lm(weight ~ bs(height, df = 5), data = women)
dat <- ggpredict(fm1, "height")

ggplot(dat, aes(x, predicted)) + 
  geom_line() +
  geom_point()
```

# Survival models

`ggpredict()` also supports `coxph`-models from the **survival**-package and is able to either plot risk-scores (the default), probabilities of survival (`type = "survival"`) or cumulative hazards (`type = "cumulative_hazard"`).

Since probabilities of survival and cumulative hazards are changing across time, the time-variable is automatically used as x-axis in such cases, so the `terms`-argument only needs up to **two** variables for `type = "survival"` or `type = "cumulative_hazard"`.

```{r}
library(survival)
data("lung2")
m <- coxph(Surv(time, status) ~ sex + age + ph.ecog, data = lung2)

# predicted risk-scores
ggpredict(m, c("sex", "ph.ecog"))
```

```{r}
# probability of survival
ggpredict(m, c("sex", "ph.ecog"), type = "survival")
```

# Labelling the data

**ggeffects** makes use of the [sjlabelled-package](https://cran.r-project.org/package=sjlabelled) and supports [labelled data](https://cran.r-project.org/package=sjlabelled/vignettes/intro_sjlabelled.html). If the data from the fitted models is labelled, the value and variable label attributes are usually copied to the model frame stored in the model object. **ggeffects** provides various _getter_-functions to access these labels, which are returned as character vector and can be used in ggplot's `lab()`- or `scale_*()`-functions.

* `get_title()` - a generic title for the plot, based on the model family, like "predicted values" or "predicted probabilities"
* `get_x_title()` - the variable label of the first model term in `terms`.
* `get_y_title()` - the variable label of the response.
* `get_legend_title()` - the variable label of the second model term in `terms`.
* `get_x_labels()` - value labels of the first model term in `terms`.
* `get_legend_labels()` - value labels of the second model term in `terms`.

The data frame returned by `ggpredict()`, `ggemmeans()` or `ggeffect()` must be used as argument to one of the above function calls.

```{r}
get_x_title(mydf)
get_y_title(mydf)

ggplot(mydf, aes(x, predicted, colour = group)) + 
  geom_line() +
  facet_wrap(~facet) +
  labs(
    x = get_x_title(mydf),
    y = get_y_title(mydf),
    colour = get_legend_title(mydf)
  )
```
