---
title: "Introduction: Adding Partial Residuals to Adjusted Predictions Plots"
author: "Daniel Lüdecke"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction: Adding Partial Residuals to Adjusted Predictions Plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r set-options, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", dev = "png", fig.width = 7, fig.height = 4, message = FALSE, warning = FALSE)
options(width = 800)
if (!requireNamespace("ggplot2", quietly = TRUE)) {
  knitr::opts_chunk$set(eval = FALSE)
}
```

Plotting partial residuals on top of the estimated marginal means allows detecting missed modeling, like unmodeled non-linear relationships or unmodeled interactions. In a nutshell, it allows _Visualizing Fit and Lack of Fit in Complex Regression Models with Predictor Effect Plots and Partial Residuals_ (Fox & Weisberg 2018).

To add partial residuals to a plot, add `show_residuals = TRUE` to the `plot()` function call. Unlike plotting raw data, partial residuals are much better in detecting spurious patterns of relationships between predictors and outcome. 

## Detecting non-linear relationship

Let's look at an example with a non-linear relationship. The missed pattern is not obvious when looking at the raw data:

```{r}
library(ggeffects)
set.seed(1234)
x <- rnorm(200)
z <- rnorm(200)
# quadratic relationship
y <- 2 * x + x^2 + 4 * z + rnorm(200)

d <- data.frame(x, y, z)
m <- lm(y ~ x + z, data = d)

pr <- predict_response(m, "x [all]")
plot(pr, show_data = TRUE)
```

However, it becomes more obvious with partial residuals:

```{r}
plot(pr, show_residuals = TRUE)
```

It is even more obvious, when a local polynomial regression line (loess) is added to the plot. This can be achieved using `show_residuals_line = TRUE`.

```{r}
plot(pr, show_residuals = TRUE, show_residuals_line = TRUE)
```

## Detecting missed interactions

Here is another example, which shows that the partial residuals plot suggests modeling an interaction:

```{r}
set.seed(1234)
x <- rnorm(300, mean = 10)
z <- rnorm(300)
v <- rnorm(300)
y <- (4 * z + 2) * x - 40 * z + 5 * v + rnorm(300, sd = 3)

d <- data.frame(x, y, z)
m <- lm(y ~ x + z, data = d)

pr <- predict_response(m, c("x", "z"))

# raw data, no interaction
plot(pr, show_data = TRUE)
```

Again, it is recommended to add a loess-fit line to the residuals:

```{r}
plot(pr, show_residuals = TRUE, grid = TRUE, show_residuals_line = TRUE)
```

Modeling the interaction clearly catches the pattern in the data better.

```{r}
m <- lm(y ~ x * z, data = d)
pr <- predict_response(m, c("x", "z"))
plot(pr, show_residuals = TRUE, grid = TRUE, show_residuals_line = TRUE)
```

## Using the complete range of values

_ggeffects_ usually "prettyfies" the data and tries to find a pretty sequence over a range of a focal predictor, to avoid too lengthy output, particularly for continuous variables (see section _pretty value ranges_ in [this vignette](introduction_effectsatvalues.html)).

This, however, might be misleading in some cases when creating residual plots. In the next example, we have a sinus-curve pattern for the residuals, which is hidden by default:

```{r}
set.seed(1234)
x <- seq(-100, 100, length.out = 1e3)
z <- rnorm(1e3)
y <- 5 * sin(x / 2) + x / 2 + 10 * z

m <- lm(y ~ x + z)
pr <- predict_response(m, "x")

plot(pr, residuals = TRUE)
```

In such cases, it is recommended to use the `all`-tag in the `terms`-argument.

```{r}
pr <- predict_response(m, "x [all]")
plot(pr, show_residuals = TRUE)
```

# References

Fox J, Weisberg S. _Visualizing Fit and Lack of Fit in Complex Regression Models with Predictor Effect Plots and Partial Residuals_. Journal of Statistical Software 2018;87. https://www.jstatsoft.org/article/view/v087i09
