---
title: "Marginal Effects of Regression Models"
author: "Daniel Lüdecke"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Marginal Effects of Regression Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r set-options, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", dev = "png", fig.width = 7, fig.height = 3.5, message = FALSE, warning = FALSE)
options(width = 800)
```

# Aim of the ggeffects-package

Results of regression models are typically presented as tables that are easy to understand. For more complex models that include interaction or quadratic / spline terms, tables with numbers are less helpful and difficult to interpret. In such cases, the visualization of _marginal effects_ is far easier to understand and allows to intuitively get the idea of how predictors and outcome are associated, even for complex models. 

The aim of this package is similar to the [broom-package](https://cran.r-project.org/package=broom): transforming "untidy" input into a tidy data frame, especially for further use with [ggplot](https://cran.r-project.org/package=ggplot2). However, **ggeffects** does not return model-summaries; rather, this package computes marginal effects at the mean or average marginal effects from statistical models and returns the result as tidy data frame.

Since the focus lies on plotting the data (the marginal effects), at least one model term needs to be specified for which the effects are computed. It is also possible to compute marginal effects for model terms, grouped by the levels of another model's predictor. The package also allows plotting marginal effects for two- or three-way-interactions, or for specific values of a model term only. Examples are shown below.

# Short technical note

`ggpredict()`, `ggeffect()`, `ggemmeans()` and `ggaverage()` always return predicted values for the _response_ of a model (or _response distribution_ for Bayesian models).

Typically, `ggpredict()` returns confidence intervals based on the standard errors (`+/- 1.96 * SE`) as returned by the `predict()`-function. If `predict()` for a certain class does _not_ return standard errors (for example, *merMod*-objects), these are calculated manually, by following steps: matrix-multiply `X` by the parameter vector `B` to get the predictions, then extract the variance-covariance matrix `V` of the parameters and compute `XVX'` to get the variance-covariance matrix of the predictions. The square-root of the  diagonal of this matrix represent the standard errors of the predictions, which are then multiplied by 1.96 for the confidence intervals.

For mixed models, if `type = "re"` or `type = "re.zi"`, the uncertainty in the random effects is accounted for when calculating the standard errors. Hence, in such cases, the intervals may be considered as _prediction intervals_.

# Consistent and tidy structure

The returned data frames always have the same, consistent structure and column names, so it's easy to create ggplot-plots without the need to re-write the arguments to be mapped in each ggplot-call. `x` and `predicted` are the values for the x- and y-axis. `conf.low` and `conf.high` could be used as `ymin` and `ymax` aesthetics for ribbons to add confidence bands to the plot. `group` can be used as grouping-aesthetics, or for faceting.

The examples shown here mostly use **ggplot2**-code for the plots, however, there is also a `plot()`-method, which is described in the vignette [Plotting Marginal Effects](plotmethod.html).

# Marginal effects at the mean

`ggpredict()` computes predicted values for all possible levels and values from a model's predictors. In the simplest case, a fitted model is passed as first argument, and the term in question as second argument. Use the raw name of the variable for the `terms`-argument only - you don't need to write things like `poly(term, 3)` or `I(term^2)` for the `terms`-argument.

```{r}
library(ggeffects)
data(efc)
fit <- lm(barthtot ~ c12hour + neg_c_7 + c161sex + c172code, data = efc)

ggpredict(fit, terms = "c12hour")
```

As you can see, `ggpredict()` (and `ggeffect()` or `ggemmeans()`) has a nice `print()`-method, which takes care of printing not too many rows (but always an equally spaced range of values, including minimum and maximum value of the term in question) and giving some extra information. This is especially useful when predicted values are shown depending on the levels of other terms (see below).

The output shows the predicted values for the response at each value from the term _c12hour_. The data is already in shape for ggplot:

```{r}
library(ggplot2)
theme_set(theme_bw())

mydf <- ggpredict(fit, terms = "c12hour")
ggplot(mydf, aes(x, predicted)) + geom_line()
```

# Marginal effects at the mean for different groups

The `terms`-argument accepts up to three model terms, where the second and third term indicate grouping levels. This allows predictions for the term in question at different levels for other model terms:

```{r}
ggpredict(fit, terms = c("c12hour", "c172code"))
```

Creating a ggplot is pretty straightforward: the colour-aesthetics is mapped with the `group`-column:

```{r}
mydf <- ggpredict(fit, terms = c("c12hour", "c172code"))
ggplot(mydf, aes(x, predicted, colour = group)) + geom_line()
```

Finally, a second grouping structure can be defined, which will create another column named `facet`, which - as the name implies - might be used to create a facted plot:

```{r}
mydf <- ggpredict(fit, terms = c("c12hour", "c172code", "c161sex"))
mydf
ggplot(mydf, aes(x, predicted, colour = group)) + 
  geom_line() + 
  facet_wrap(~facet)
```

# Marginal effects for each model term

If the `term` argument is either missing or `NULL`, marginal effects for each model term are calculated. The result is returned as a list, which can be plotted manually (or using the `plot()` function).

```{r}
mydf <- ggpredict(fit)
mydf
```


# Average marginal effects

`ggaverage()` compute average marginal effects. While `ggpredict()` creates a data-grid (using `expand.grid()`) for all possible combinations of values (even if some combinations are not present in the data), `ggaverage()` computes predicted values based on the given data. This means that _different_ predicted values for the outcome may occure at the _same_ value or level for the term in question. The predicted values are then averaged for each value of the term in question and the linear trend is smoothed accross the averaged predicted values. This means that the line representing the marginal effects may cross or diverge, and are not necessarily in paralell to each other.

```{r}
mydf <- ggaverage(fit, terms = c("c12hour", "c172code"))
ggplot(mydf, aes(x, predicted, colour = group)) + geom_line()
```

Since `ggaverage()` produces _different_ predicted values for the _same_ value or level of terms, linear-smoothing for gaussian models and loess-smoothing for non-gaussian models is used to produce "smooth" lines. This may lead to misleading plots, especially if polynomial or spline terms are included in the model. It's preferred to use `ggpredict()` or `ggeffect()` resp. `ggemmeans()`.

# Two- and Three-Way-Interactions

To plot the marginal effects of interaction terms, simply specify these terms in the `terms`-argument.

```{r}
library(sjmisc)
data(efc)

# make categorical
efc$c161sex <- to_factor(efc$c161sex)

# fit model with interaction
fit <- lm(neg_c_7 ~ c12hour + barthtot * c161sex, data = efc)

# select only levels 30, 50 and 70 from continuous variable Barthel-Index
mydf <- ggpredict(fit, terms = c("barthtot [30,50,70]", "c161sex"))
ggplot(mydf, aes(x, predicted, colour = group)) + geom_line()
```

Since the `terms`-argument accepts up to three model terms, you can also compute marginal effects for a 3-way-interaction. To plot the marginal effects of three interaction terms, just like before, specify all three terms in the `terms`-argument.

```{r}
# fit model with 3-way-interaction
fit <- lm(neg_c_7 ~ c12hour * barthtot * c161sex, data = efc)

# select only levels 30, 50 and 70 from continuous variable Barthel-Index
mydf <- ggpredict(fit, terms = c("c12hour", "barthtot [30,50,70]", "c161sex"))

ggplot(mydf, aes(x, predicted, colour = group)) + 
  geom_line() +
  facet_wrap(~facet)
```

# Polynomial terms and splines

`ggpredict()` also works for models with polynomial terms or splines. Following code reproduces the plot from `?splines::bs`:


```{r}
library(splines)
data(women)

fm1 <- lm(weight ~ bs(height, df = 5), data = women)
dat <- ggpredict(fm1, "height")

ggplot(dat, aes(x, predicted)) + 
  geom_line() +
  geom_point()
```

# Survival models

`ggpredict()` also supports `coxph`-models from the **survival**-package and is able to either plot risk-scores (the default), probabilities of survival (`type = "surv"`) or cumulative hazards (`type = "cumhaz"`).

Since probabilities of survival and cumulative hazards are changing accross time, the time-variable is automatically used as x-axis in such cases, so the `terms`-argument only needs up to **two** variables for `type = "surv"` or `type = "cumhaz"`.

```{r}
data("lung", package = "survival")
# remove category 3 (outlier)
lung <- subset(lung, subset = ph.ecog %in% 0:2)
lung$sex <- factor(lung$sex, labels = c("male", "female"))
lung$ph.ecog <- factor(lung$ph.ecog, labels = c("good", "ok", "limited"))

m <- survival::coxph(survival::Surv(time, status) ~ sex + age + ph.ecog, data = lung)

# predicted risk-scores
ggpredict(m, c("sex", "ph.ecog"))
```

```{r}
# probability of survival
ggpredict(m, c("sex", "ph.ecog"), type = "surv")
```

# Printing factor levels instead of numeric values

By default, to quickly create plot where data points can be connected with lines (i.e. to use a continuous x-axis-scale), the `x`-value is numeric, even if the first variable in `terms` (which is represented by the `x`-column in the returned data frame) is categorical. Since **ggeffects** supports labelled data (see also next section), plots are always correctly annotated, and a discrete scale is automatically used (see vignette [Plotting Marginal Effects](plotmethod.html)).

If you want to preserve the factor-class for `x`, use the argument `x.as.factor = TRUE`. This will also print the factor levels instead of numeric values in the console. An alternative would be to use the  `x.lab = TRUE`-argument in the `print()`-method.

Here are some examples, first we fit a model where two variables are factors with character factor levels (we use `sjmisc::to_label()` to convert the labelled data into factors):

```{r}
data(efc)
efc <- to_label(efc, c172code, c161sex)
m <- lm(barthtot ~ c12hour + neg_c_7 + c161sex + c172code, data = efc)
pr <- ggpredict(m, "c172code")
```

By default, `x` is numeric.

```{r}
print(pr)
```

As value labels and factor levels are saved as attributes in the returned data frame, we still can print these value labels:

```{r}
print(pr, x.lab = TRUE)
```

Again, printing the results will cause factors as `x` to become numeric...

```{r}
ggpredict(m, "c172code")
```

... unless you change this behaviour with `x.as.factor = TRUE`.

```{r}
ggpredict(m, "c172code", x.as.factor = TRUE)
```

# Labelling the data

**ggeffects** makes use of the [sjlabelled-package](https://cran.r-project.org/package=sjlabelled) and supports [labelled data](https://cran.r-project.org/package=sjlabelled/vignettes/intro_sjlabelled.html). If the data from the fitted models is labelled, the value and variable label attributes are usually copied to the model frame stored in the model object. **ggeffects** provides various _getter_-functions to access these labels, which are returned as character vector and can be used in ggplot's `lab()`- or `scale_*()`-functions.

* `get_title()` - a generic title for the plot, based on the model family, like "predicted values" or "predicted probabilities"
* `get_x_title()` - the variable label of the first model term in `terms`.
* `get_y_title()` - the variable label of the response.
* `get_legend_title()` - the variable label of the second model term in `terms`.
* `get_x_labels()` - value labels of the first model term in `terms`.
* `get_legend_labels()` - value labels of the second model term in `terms`.

The data frame returned by `ggpredict()` or `ggaverage()` must be used as argument to one of the above function calls.

```{r}
get_x_title(mydf)
get_y_title(mydf)

ggplot(mydf, aes(x, predicted, colour = group)) + 
  geom_line() +
  facet_wrap(~facet) +
  labs(
    x = get_x_title(mydf),
    y = get_y_title(mydf),
    colour = get_legend_title(mydf)
  )
```
