---
title: "Plotting Marginal Effects"
author: "Daniel Lüdecke"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plotting Marginal Effects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r set-options, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", dev = "png", fig.width = 7, fig.height = 4, message = FALSE, warning = FALSE)
options(width = 800)
```

# plot()-method

This vignettes demonstrates the `plot()`-method of the **ggeffects**-package. It is recommended to read the [general introduction](marginaleffects.html) first, if you haven't done this yet.

If you don't want to write your own ggplot-code, **ggeffects** has a `plot()`-method with some convenient defaults, which allows quickly creating ggplot-objects. `plot()` has some arguments to tweak the plot-appearance. For instance, `ci` allows you to show or hide confidence bands (or error bars, for discrete variables), `facets` allows you to create facets even for just one grouping variable, or `colors` allows you to quickly choose from some color-palettes, including black & white colored plots. Use `rawdata` to add the raw data points to the plot.

```{r}
library(ggeffects)
library(sjmisc)
data(efc)
efc$c172code <- to_label(efc$c172code)
fit <- lm(barthtot ~ c12hour + neg_c_7 + c161sex + c172code, data = efc)

# facet by group
dat <- ggpredict(fit, terms = c("c12hour", "c172code"))
plot(dat, facet = TRUE)
# don't use facets, b/w figure, w/o confidence bands
plot(dat, colors = "bw", ci = FALSE)
# plot raw data
dat <- ggpredict(fit, terms = c("c12hour", "c172code"))
plot(dat, rawdata = TRUE)
# for three variables, automatic facetting
dat <- ggpredict(fit, terms = c("c12hour", "c172code", "c161sex"))
plot(dat)
# categorical variables have errorbars
dat <- ggpredict(fit, terms = c("c172code", "c161sex"))
plot(dat)
```

# Change appearance of confidence bands

In some plots, the the confidence bands are not represented by a shaded area (ribbons), but rather by dashed or dotted lines. Use `ci = "dash"` or `ci = "dot"` to change the style of confidence bands.

```{r}
# dashed lines for CI
dat <- ggpredict(fit, terms = "c12hour")
plot(dat, ci = "dash")
# facet by group
dat <- ggpredict(fit, terms = c("c12hour", "c172code"))
plot(dat, facet = TRUE, ci = "dot")
```

# Log-transform y-axis for binomial models

For binomial models, the y-axis indicates the predicted probabilities of an event. In this case, error bars are not symmetrical.

```{r}
if (requireNamespace("lme4")) {
  m <- glm(
    cbind(incidence, size - incidence) ~ period, 
    family = binomial, 
    data = lme4::cbpp
  )
  
  dat <- ggpredict(m, "period")
  
  # normal plot, asymmetrical error bars
  plot(dat)
}
```

Here you can use `log.y` to log-transform the y-axis. The `plot()`-method will automatically choose axis breaks and limits that fit well to the value range and log-scale.
 
```{r}
if (requireNamespace("lme4")) {
  # plot with log-transformed y-axis
  plot(dat, log.y = TRUE)
}
```

# Control y-axis appearance

Furthermore, arguments in `...` are passed down to `ggplot::scale_y_continuous()` (resp. `ggplot::scale_y_log10()`, if `log.y = TRUE`), so you can control the appearance of the y-axis.

```{r}
if (requireNamespace("lme4")) {
  # plot with log-transformed y-axis, modify breaks
  plot(
    dat, log.y = TRUE, 
    breaks = c(.05, .1, .15, .2, .25, .3),
    limits = c(.01, .3)
  )
}
```
