---
title: "Case Study: Predictions for Mixed Models: Comparison of {ggeffects} and {marginaleffects}"
author: "Daniel Lüdecke"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Case Study: Predictions for Mixed Models: Comparison of ggeffects and marginaleffects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r set-options, echo=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", dev = "png", fig.width = 7, fig.height = 4, message = FALSE, warning = FALSE)
options(width = 800)
if (!requireNamespace("ggplot2", quietly = TRUE) ||
    !requireNamespace("lme4", quietly = TRUE) ||
    !requireNamespace("patchwork", quietly = TRUE) ||
    !requireNamespace("marginaleffects", quietly = TRUE) ||
    !requireNamespace("sjlabelled", quietly = TRUE)) {
  knitr::opts_chunk$set(eval = FALSE)
}
```

This vignette compares the *ggeffects* package with the *marginaleffects* package, which is another package that can be used to create predictions for mixed models. It shows how to reproduce the plots shown in the [mixed models case study for population- and unit-level predictions](https://marginaleffects.com/vignettes/lme4.html)

It is probably a good idea to read the [introduction to predictions for mixed models](https://strengejacke.github.io/ggeffects/articles/introduction_randomeffects.html) first, to get familiar with the basics of the package regarding adjusted predictions for mixed models.

## Population-level and unit-level predictions: comparison between {marginaleffects} and {ggeffects}

First, we fit the two example models.

```{r}
library(lme4)
library(patchwork)
library(marginaleffects)
library(ggplot2)

data(ChickWeight)

model1 <- lmer(
  weight ~ 1 + Time + (1 + Time | Chick),
  data = ChickWeight
)

model2 <- lmer(
  weight ~ 1 + Time + I(Time^2) + Diet + Time:Diet + I(Time^2):Diet +
    (1 + Time + I(Time^2) | Chick),
  data = ChickWeight
)
```

### Unit-level predictions

The first two plots show unit-level predictions created with the *marginaleffects* package. As can be seen, predictions for each level of the random effects are included.

```{r}
pred1 <- predictions(model1,
  newdata = datagrid(
    Chick = ChickWeight$Chick,
    Time = 0:21
  )
)

p1 <- ggplot(pred1, aes(Time, estimate, level = Chick)) +
  geom_line() +
  labs(y = "Predicted weight", x = "Time", title = "Linear growth model")

pred2 <- predictions(model2,
  newdata = datagrid(
    Chick = ChickWeight$Chick,
    Time = 0:21
  )
)

p2 <- ggplot(pred2, aes(Time, estimate, level = Chick)) +
  geom_line() +
  labs(y = "Predicted weight", x = "Time", title = "Quadratic growth model")

p1 + p2
```

*ggeffects* handles unit-level predictions slightly different (see [this vignette](https://strengejacke.github.io/ggeffects/articles/introduction_randomeffects.html)) - each unit-level is considered as own "group", thus the plot would normally use colors and a color legend to distinguish between the unit-levels. In this example, the default color palette is too small to plot all unit-levels.

```{r eval=FALSE}
pr <- ggpredict(fit1, terms = c("Time", "Chick"), type = "random")
plot(pr)
#> Error in `palette()`:
#>
```

To reproduce the plots from the *marginaleffects* package, we need to modify our plot. We simply provide a vector with a sufficient amount of color values and hide the legend.

```{r}
pr <- ggpredict(model1, terms = c("Time", "Chick"), type = "random")
p3 <- plot(pr, colors = rep("grey30", 50), show_ci = FALSE, show_legend = FALSE)

pr <- ggpredict(model2, terms = c("Time", "Chick"), type = "random")
p4 <- plot(pr, colors = rep("grey30", 50), show_ci = FALSE, show_legend = FALSE)

p3 + p4
```

### Unit-level predictions stratified by `Diet`

This is the next plot shown in the *marginaleffects* case study. Unit-level predictions are stratified by `Diet`.

```{r}
pred <- predictions(model2)

ggplot(pred, aes(Time, estimate, level = Chick)) +
  geom_line() +
  ylab("Predicted Weight") +
  facet_wrap(~Diet, labeller = label_both)
```

*ggeffects* by default creates a (theoretical) reference grid for all possible combinations in the data. That's why the following plot looks different than the one above. We see predictions for *all* unit-levels in each panel of `Diet`.

```{r, warning=FALSE, message=FALSE}
pr <- ggpredict(model2, terms = c("Time", "Chick", "Diet"), type = "random")
plot(pr, colors = rep("grey30", 50), show_ci = FALSE, show_legend = FALSE)
```

To limit the plot to the unit-levels that are actually present in the data, we need to set `limit_range = TRUE`.

```{r, warning=FALSE, message=FALSE}
pr <- ggpredict(model2, terms = c("Time", "Chick", "Diet"), type = "random")
plot(
  pr,
  colors = rep("grey30", 50),
  show_ci = FALSE,
  show_legend = FALSE,
  limit_range = TRUE
)
```

### Population-level predictions

This last example show population-level predictions.

```{r}
pred <- predictions(
  model2,
  newdata = datagrid(
    Chick = NA,
    Diet = 1:4,
    Time = 0:21
  ),
  re.form = NA
)

ggplot(pred, aes(x = Time, y = estimate, ymin = conf.low, ymax = conf.high)) +
  geom_ribbon(alpha = 0.1, fill = "red") +
  geom_line() +
  facet_wrap(~Diet, labeller = label_both) +
  labs(title = "Population-level trajectories")
```

This plot is rather simple to reproduce with *ggeffects*. We don't need to specify the `type` argument, since `type = "fixed"` is the default and returns population-level predictions.

```{r}
pr <- ggpredict(fit2, terms = c("Time", "Diet"))
plot(pr, grid = TRUE)
```
