---
title: "Marginal Effects for Random Effects"
author: "Daniel Lüdecke"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Marginal Effects for Random Effects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r set-options, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", dev = "png", fig.width = 7, fig.height = 4, message = FALSE, warning = FALSE)
options(width = 800)
```

## Marginal effects for each level of random effects

Marginal effects can also be calculated for each group level in mixed models. Simply add the name of the related random effects term to the `terms`-argument, and set `type = "re"`.

In the following example, we fit a linear mixed model and first simply plot the marginal effetcs, _not_ conditioned on random effects.

```{r}
library(sjlabelled)
library(lme4)
data(efc)
efc$e15relat <- as_label(efc$e15relat)
m <- lmer(neg_c_7 ~ c12hour + c160age + c161sex + (1 | e15relat), data = efc)
me <- ggpredict(m, terms = "c12hour")
plot(me)
```

Changing the type to `type = "re"` still returns population-level predictions by default. The major difference between `type = "fe"` and `type = "re"` is the uncertainty in the variance parameters. This leads to larger confidence intervals for marginal effects with `type = "re"`.

```{r}
me <- ggpredict(m, terms = c("c12hour"), type = "re")
plot(me)
```

To compute marginal effects for each grouping level, add the related random term to the `terms`-argument. In this case, confidence intervals are not calculated, but marginal effects are conditioned on each group level of the random effects.

```{r}
me <- ggpredict(m, terms = c("c12hour", "e15relat"), type = "re")
plot(me)
```

Marginal effects, conditioned on random effects, can also be calculated for specific levels only. Add the related values into brackets after the variable name in the `terms`-argument.

```{r}
me <- ggpredict(m, terms = c("c12hour", "e15relat [child,sibling]"), type = "re")
plot(me)
```

If the group factor has too many levels, you can also take a random sample of all possible levels and plot the marginal effects for this subsample of group levels. To do this, use `term = "<groupfactor> [sample=n]"`.

```{r}
data("sleepstudy")
m <- lmer(Reaction ~ Days + (1 + Days | Subject), data = sleepstudy)
me <- ggpredict(m, terms = c("Days", "Subject [sample=8]"), type = "re")
plot(me)
```
