---
title: "Technical Details: Difference between marginalization methods: the `margin` argument"
author: "Daniel Lüdecke"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Technical Details: Difference between marginalization methods: the `margin` argument}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r set-options, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", dev = "png", fig.width = 7, fig.height = 3.5, message = FALSE, warning = FALSE)
options(width = 800)
if (!requireNamespace("magrittr", quietly = TRUE) ||
    !requireNamespace("see", quietly = TRUE) ||
    !requireNamespace("emmeans", quietly = TRUE) ||
    !requireNamespace("marginaleffects", quietly = TRUE) ||
    !requireNamespace("datawizard", quietly = TRUE)) {
  knitr::opts_chunk$set(eval = FALSE)
}
```

`predict_response()` computes marginal means or predicted values for all possible levels or values from specified model's predictors (_focal terms_). These effects are "marginalized" (or "averaged") over the values or levels of remaining predictors (the **non-focal** terms). The `margin` argument specifies the method of marginalization. The following methods are available:

- `"mean_reference"`: non-focal predictors are set to their mean (numeric variables), reference level (factors), or "most common" value (mode) in case of character vectors. Technically, a data grid is constructed, roughly comparable to `expand.grid()` on all unique combinations of `model.frame(model)[, terms]`. This data grid (see `data_grid()`) is used for the `newdata` argument of `predict()`. All remaining covariates not specified in `terms` are held constant: Numeric values are set to the mean, integer values are set to their median, factors are set to their reference level and character vectors to their mode (most common element).

- `"mean_mode"`: non-focal predictors are set to their mean (numeric variables) or mode (factors, or "most common" value in case of character vectors).

- `"marginalmeans"`: non-focal predictors are set to their mean (numeric variables) or marginalized over the levels or "values" for factors and character vectors. Marginalizing over the factor levels of non-focal terms computes a kind of "weighted average" for the values at which these terms are hold constant.

- `"empirical"` (or `"counterfactual"`): non-focal predictors are marginalized over the observations in your sample, for _counterfactual predictions_. Technically, predicted values for each observation in the data are calculated multiple times (the data is duplicated once for all unique values of the focal terms), each time fixing one unique value or level of the focal terms and then takes the average of these predicted values (aggregated/grouped by the focal terms).

That means:

- For models without categorical predictors, results are usually identical, no matter which `margin` option is selected (except some _slight_ differences in the associated confidence intervals, which are, however, negligible).

- When all categorical predictors are specified in `terms` and further (non-focal) terms are only numeric, results are usually identical, as well.

```{r}
library(magrittr)
library(ggeffects)
data(efc, package = "ggeffects")
fit <- lm(barthtot ~ c12hour + neg_c_7, data = efc)

# we add margin = "mean_reference" to show that it is the default
predict_response(fit, "c12hour", margin = "mean_reference")

predict_response(fit, "c12hour", margin = "marginalmeans")
```

As can be seen, the continuous predictor `neg_c_7` is held constant at its mean value, 11.83. For categorical predictors, `margin = "mean_reference"` (the default, and thus not specified in the above example) and `margin = "marginalmeans"` behave differently. While `"mean_reference"` uses the reference level of each categorical predictor to hold it constant, `"marginalmeans"` averages over the proportions of the categories of factors.

```{r}
library(datawizard)
data(efc, package = "ggeffects")
efc$e42dep <- to_factor(efc$e42dep)
# we add categorical predictors to our model
fit <- lm(barthtot ~ c12hour + neg_c_7 + e42dep, data = efc)

predict_response(fit, "c12hour", margin = "mean_reference")

predict_response(fit, "c12hour", margin = "marginalmeans")
```

In this case, one would obtain the same results for `"mean_reference"` and `"marginalmeans"` again, if `condition` is used to define specific levels at which variables, in our case the factor `e42dep`, should be held constant.

```{r}
predict_response(fit, "c12hour", margin = "mean_reference")

predict_response(
  fit,
  "c12hour",
  margin = "marginalmeans",
  condition = c(e42dep = "independent")
)
```

Another option is to use `predict_response(margin = "empirical")` to compute "counterfactual" adjusted predictions. This function is a wrapper for the `avg_predictions()`-method from the **marginaleffects**-package. The major difference to `margin = "marginalmeans"` is that estimated marginal means, as computed by `"marginalmeans"`, are a special case of predictions, made on a perfectly balanced grid of categorical predictors, with numeric predictors held at their means, and marginalized with respect to some focal variables. `predict_response(margin = "empirical")`, in turn, calculates predicted values for each observation in the data multiple times, each time fixing the unique values or levels of the focal terms to one specific value and then takes the average of these predicted values (aggregated/grouped by the focal terms) - or in other words: the whole dataset is duplicated once for every unique value of the focal terms, makes predictions for each observation of the new dataset and take the average of all predictions (grouped by focal terms). This is also called "counterfactual" predictions.

```{r}
predict_response(fit, "c12hour", margin = "empirical")
```

To explain how `margin = "empirical"` works, let's look at following example. The confidence intervals for the simple means differ from those of `predict_response()`, however, the predicted and mean values are identical.

```{r}
data(iris)
set.seed(123)
# create an unequal distributed factor, used as focal term
iris$x <- as.factor(sample(1:4, nrow(iris), replace = TRUE, prob = c(0.1, 0.2, 0.3, 0.4)))
m <- lm(Sepal.Width ~ Species + x, data = iris)

# average predicted values
predict_response(m, "x", margin = "empirical")

# replicate the dataset for each level of "x", i.e. 4 times
d <- do.call(rbind, replicate(4, iris, simplify = FALSE))
# for each data set, we set our focal term "x" to one of the four unique values
d$x <- as.factor(rep(1:4, each = 150))
# we calculate predicted values for each "dataset", i.e. we predict our outcome
# for observations, for all levels of "x"
d$predicted <- predict(m, newdata = d)
# now we compute the average predicted values for each level of "x"
datawizard::means_by_group(d, "predicted", "x")
```

**But when should I use which `margin` option?**

When you are interested in the strength of association, it usually doesn't matter. as you can see in the plots below. The slope of our focal term, `c12hour`, is the same for all four plots:

```{r}
library(see)
predicted_1 <- predict_response(fit, terms = "c12hour")
predicted_2 <- predict_response(fit, terms = "c12hour", margin = "marginalmeans")
predicted_3 <- predict_response(fit, terms = "c12hour", margin = "marginalmeans", condition = c(e42dep = "independent"))
predicted_4 <- predict_response(fit, terms = "c12hour", margin = "empirical")

p1 <- plot(predicted_1, show_ci = FALSE, show_title = FALSE, show_x_title = FALSE, show_y_title = FALSE)
p2 <- plot(predicted_2, show_ci = FALSE, show_title = FALSE, show_x_title = FALSE, show_y_title = FALSE)
p3 <- plot(predicted_3, show_ci = FALSE, show_title = FALSE, show_x_title = FALSE, show_y_title = FALSE)
p4 <- plot(predicted_4, show_ci = FALSE, show_title = FALSE, show_x_title = FALSE, show_y_title = FALSE)

plots(p1, p2, p3, p4, n_rows = 2)
```

However, the predicted outcome varies. This gives an impression when `margin = "marginalmeans"`, i.e. _marginal_ effects, matter: when you want to predict your outcome, marginalized over the different levels of factors, i.e. "generalized" to the population (of your sample) on a _balanced grid_ of focal terms. `margin = "mean_reference"` would give a predicted outcome for a _subgroup_ (or: specific group) of your sample, i.e. conditioned on specific levels of factors. Hence, the predicted outcome does not necessarily generalize to the "population" (always keeping in mind that we assume having a "representative sample" of a "population" as data in our model). Finally, `margin = "empirical"` gives you predictions averaged across your sample and aggregated by the focal terms, and probably reflects your data most realistically.

**What is the most apparent difference from `margin = "empirical"` to the other options?**

The most apparent difference from `margin = "empirical"` compared to the other methods occurs when you have categorical co-variates (*non-focal terms*) with unequally distributed levels. `margin = "marginalmeans"` will "average" over the levels of non-focal factors, while `margin = "empirical"` will average over the observations in your sample.

Let's show this with a very simple example:

```{r}
data(iris)
set.seed(123)
# create an unequal distributed factor, used as non-focal term
iris$x <- as.factor(sample(1:4, nrow(iris), replace = TRUE, prob = c(0.1, 0.2, 0.3, 0.4)))
m <- lm(Sepal.Width ~ Species + x, data = iris)

# predicted values, conditioned on x = 1
predict_response(m, "Species")

# predicted values, conditioned on weighted average of x
predict_response(m, "Species", margin = "marginalmeans")

# average predicted values, averaged over the sample and aggregated by "Species"
predict_response(m, "Species", margin = "empirical")
```

There is no rule of thumb which approach is better; it depends on the characteristics of the sample and the population to which should be generalized. Consulting the [marginaleffects-website](https://marginaleffects.com/) might help to decide which approach is more appropriate.

**Showing the difference between `"marginalmeans"` and `"empirical"`**

A code example, where we compute the average predicted values the estimated marginal means manually, shows the differences between these two ways of averaging in detail.

```{r}
data(iris)
set.seed(123)
iris$x <- as.factor(sample(1:4, nrow(iris), replace = TRUE, prob = c(0.1, 0.2, 0.3, 0.4)))
m <- lm(Sepal.Width ~ Species + x, data = iris)

# average predicted values
predict_response(m, "Species", margin = "empirical")

# replicate the dataset for each level of "Species", i.e. 3 times
d <- do.call(rbind, replicate(3, iris, simplify = FALSE))
# for each data set, we set our focal term to one of the three levels
d$Species <- as.factor(rep(levels(iris$Species), each = 150))
# we calculate predicted values for each "dataset", i.e. we predict our outcome
# for observations, for all levels of "Species"
d$predicted <- predict(m, newdata = d)
# now we compute the average predicted values by levels of "Species"
datawizard::means_by_group(d, "predicted", "Species")

# estimated marginal means, in turn, differ from the above, because they are
# averaged across balanced reference grids for all focal terms, thereby non-focal
# are hold constant at a "weighted average".

# estimated marginal means, from `ggemmeans()`
predict_response(m, "Species", margin = "marginalmeans")

d <- rbind(
  data_grid(m, "Species", condition = c(x = "1")),
  data_grid(m, "Species", condition = c(x = "2")),
  data_grid(m, "Species", condition = c(x = "3")),
  data_grid(m, "Species", condition = c(x = "4"))
)
d$predicted <- predict(m, newdata = d)
# means calculated manually
datawizard::means_by_group(d, "predicted", "Species")
```

**When should I use which `margin` option?**

The *strength* of the association should not be affected by the way of averaging. You can see this when comparing the differences between the predicted values, which are the same. If you're interested in differences between specific levels (or values) of your predictors, it shouldn't matter which function you use.

What _is_ different is the predicted outcome, and therefore the conclusions you would draw from the results. `margin = "mean_reference"` or `margin = "mean_mode"` predict your outcome for certain characteristics of your sample (for "subgroups", if you like). `margin = "empirical"` and `margin = "marginalmeans"` predict your outcome for an "average observation" of your sample, with slightly different approaches of averaging.

```{r}
# check that the difference between the predicted value, average predicted values
#  estimated marginal means is the same
out1 <- predict_response(m, "Species", margin = "empirical")
out2 <- predict_response(m, "Species", margin = "marginalmeans")
out3 <- predict_response(m, "Species", margin = "mean_reference")
all.equal(diff(out1$predicted), diff(out2$predicted))
all.equal(diff(out1$predicted), diff(out3$predicted))
```

Some models are not yet supported by the **emmeans** package, thus, for certain models, only `"mean_reference"` works, not `"marginalmeans"`. `"empirical"` should support the same (or more) models as `"mean_reference"`, but averages predictions from the sample, and thus may not be the quantity of interest. Sometimes, robust variance-covariance estimation is required for confidence intervals of predictions. In such cases, you have to rely on `"mean_reference"` or `"empirical"`. If you have no categorical predictors as non-focal terms (i.e. no factor needs to be held constant), then - as shown above - `"mean_reference"` and `"marginalmeans"` yield the same results.
