---
title: "Technical Details: Difference between ggpredict() and ggemmeans()"
author: "Daniel Lüdecke"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Technical Details: Difference between ggpredict() and ggemmeans()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r set-options, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", dev = "png", fig.width = 7, fig.height = 3.5, message = FALSE, warning = FALSE)
options(width = 800)
if (!requireNamespace("magrittr", quietly = TRUE) ||
    !requireNamespace("see", quietly = TRUE) ||
    !requireNamespace("sjmisc", quietly = TRUE)) {
  knitr::opts_chunk$set(eval = FALSE)
}
```

`ggpredict()` and `ggemmeans()` compute predicted values for all possible levels or values from a model's predictor. Basically, `ggpredict()` wraps the `predict()`-method for the related model, while `ggemmeans()` wraps the `emmeans()`-method from the **emmeans**-package. Both `ggpredict()` and `ggemmeans()` do some data-preparation to bring the data in shape for the `newdata`-argument (`predict()`) resp. the `at`-argument (`emmeans()`). It is recommended to read the [general introduction](ggeffects.html) first, if you haven't done this yet.

Thus, effects returned by `ggpredict()` _conditional effects_ (i.e. these are conditioned on certain (reference) levels of factors), while `ggemmeans()` returns _marginal means_, since the effects are "marginalized" (or "averaged") over the levels of factors.

For models without categorical predictors, the results from `ggpredict()` and `ggemmeans()` are identical (except some _slight_ differences in the associated confidence intervals, which are, however, negligible).

```{r}
library(magrittr)
library(ggeffects)
data(efc)
fit <- lm(barthtot ~ c12hour + neg_c_7, data = efc)

ggpredict(fit, terms = "c12hour")

ggemmeans(fit, terms = "c12hour")
```

As can be seen, the continuous predictor `neg_c_7` is held constant at its mean value, 11.83. For categorical predictors, `ggpredict()` and `ggemmeans()` behave differently. While `ggpredict()` uses the reference level of each categorical predictor to hold it constant, `ggemmeans()` - like `ggeffect()` - averages over the proportions of the categories of factors.

```{r}
library(sjmisc)
data(efc)
efc$e42dep <- to_label(efc$e42dep)
fit <- lm(barthtot ~ c12hour + neg_c_7 + e42dep, data = efc)

ggpredict(fit, terms = "c12hour")

ggemmeans(fit, terms = "c12hour")
```

In this case, one would obtain the same results for `ggpredict()` and `ggemmeans()` again, if `condition` is used to define specific levels at which variables, in our case the factor `e42dep`, should be held constant.

```{r}
ggpredict(fit, terms = "c12hour")

ggemmeans(fit, terms = "c12hour", condition = c(e42dep = "independent"))
```

Creating plots is as simple as described in the vignette [Plotting Marginal Effects](introduction_plotmethod.html).

```{r}
ggemmeans(fit, terms = c("c12hour", "e42dep")) %>% plot()
```

**But when should I use `ggemmeans()` and when `ggpredict()`?**

When you are interested in the strength of association, it usually doesn't matter. as you can see in the plots below. The slope of our focal term, `c12hour`, is the same for all three plots:

```{r}
library(see)
p1 <- plot(ggpredict(fit, terms = "c12hour"), ci = FALSE, show.title = FALSE, show.x.title = FALSE, show.y.title = FALSE)
p2 <- plot(ggemmeans(fit, terms = "c12hour"), ci = FALSE, show.title = FALSE, show.x.title = FALSE, show.y.title = FALSE)
p3 <- plot(ggemmeans(fit, terms = "c12hour", condition = c(e42dep = "independent")), ci = FALSE, show.title = FALSE, show.x.title = FALSE, show.y.title = FALSE)

plots(p1, p2, p3, n_rows = 1)
```

However, the predicted outcome varies. This gives an impression when `ggemmeans()`, i.e. _marginal_ effects, matter: when you want to predict your outcome, marginalized over the different levels of factors, i.e. "generalized" to the population (of your sample). `ggpredict()` would give a predicted outcome for a _subgroup_ of your sample, i.e. conditioned on specific levels of factors. Hence, the predicted outcome from `ggpredict()` does not necessarily generalize to the "population" (always keeping in mind that we assume having a "representative sample" of a "population" as data in our model).
